### 2020-04-05 06-examples.tex
small adjustments
modificati 7 files
modificati 7 files

Here, for the entities $q$ and $w$ that are not provided by the context, we let:
\[
\begin{array}{lcl@{\hspace{1cm}}lcl}
P_q & \defeq & E(q,\widetilde{q}) & \overline{P}_q & \defeq & E(\overline{q},\widetilde{q}) \\
P_w& \defeq & E(w,\widetilde{w}) & \overline{P}_w & \defeq & E(\overline{w},\widetilde{w})
\end{array}
\]
where:
\[
\begin{array}{lcl}
E(q,\widetilde{q})&\defeq & \sum_{h=1}^3 (\startchain{q_i}\chainedlink{q_o}{\noact}\chainend{\noact})^h\ \link{\widetilde{q}_i}{\widetilde{q}_o}.P_q\\
&& +\\
&& \sum_{h=1}^3 (\startchain{q_i}\chainedlink{q_o}{\noact}\chainend{\noact})^h.\overline{P}_q\\
\end{array}
\]
In fact the presence/absence of $q$ and $w$ will be exploited by at least one rule and as most three rules.

@@

Now, we assume that we have an initial configuration containing the entities $q$ and $b$:
% (with $\widetilde{q},\widetilde{w},\widetilde{a},\widetilde{b}$ grouping all the possibile decorations for these names):
\[
\begin{array}{lcl}
\mathit{Sys}  \defeq \ %\restrict{r1,r2,r3,r4,\widetilde{q},\widetilde{w},\widetilde{a},\widetilde{b}}
 I\mid P_q \mid \overline{P}_w  \mid  \overline{P}_a  \mid P_b  \mid  P_1  \mid  P_2  \mid  P_3  \mid P_4  \mid \mathit{Cxt}.
\end{array}
\] 
Then, only the second reaction can be applied, and the transition carries the complete label $\upsilon$ below
{\tiny
\[
%\begin{array}{l}
%\upsilon = \\
%\restrict{r1,r2,r3,r4,\widetilde{q},\widetilde{w},\widetilde{a},\widetilde{b}}\\
 \startchain{\silent}
 \chainedlink{r_1}{r_1}
 \chainedlink{\overline{a}_i}{\bf  \overline{a}_i}\chainedlink{\bf \overline{a}_o}{\overline{a}_o}
 \chainedlink{r_2}{r_2}
 \chainedlink{q_i}{\bf q_i}\chainedlink{\bf q_o}{q_o}
 \chainedlink{b_i}{\bf b_i}\chainedlink{\bf b_o}{b_io}
 \chainedlink{\overline{w}_i}{\bf \overline{w}_i}\chainedlink{\bf \overline{w}_o}{\overline{w}_o}
 \chainedlink{\overline{a}_i}{\bf \overline{a}_i}\chainedlink{\bf \overline{a}_o}{\overline{a}_o}
 \chainedlink{r_3}{r_3}
 \chainedlink{\overline{a}_i}{\bf \overline{a}_i}\chainedlink{\bf \overline{a}_o}{\overline{a}_o}
 \chainedlink{r_4}{r_4}
 \chainedlink{\overline{w}_i}{\bf \overline{w}_i}\chainedlink{\bf \overline{w}_o}{\overline{w}_o}
 \chainedlink{\mathit{cxt}}{\mathit{cxt}}
 \chainedlink{\widehat{a}_i}{\bf \widehat{a}_i}\chainedlink{\bf \widehat{a}_o}{\widehat{a}_o}
 \chainedlink{\underline{b}_i}{\bf \underline{b}_i}\chainedlink{\bf \underline{b}_o}{\underline{b}_o}
 \chainedlink{p_1}{p_1}
 \chainedlink{p_2}{p_2}
 \chainedlink{\widetilde{w}_i}{\bf \widetilde{w}_i}\chainedlink{\bf \widetilde{w}_o}{\widetilde{w}_o}
 \chainedlink{p_3}{p_3}
 \chainedlink{p_4}{p_4}
 \chainend{\silent}
%\end{array}
\]}
The parts in bold are provided by the entity processes, the other parts are provided by the processes encoding the reactions and by the process encoding the context (starting at $\mathit{cxt}$ and ended at $p_1$.
In the label we can read that the rules $1$ and $4$ have been not executed because the entity $a$ is absent, 
the rule $3$ has been not applied because the entity $w$ is absent, then only rule $2$ has been applied, and it has produced the entity $w$. Also,  the context provides entity $a$, that will be available in the next state, and not the entity $b$.
Now, to let the label more readable, we show the result of the application of the function $flat(\cdot)$ to it:
\[
r_1~\overline{a}~r_2~q~b~\overline{w}~\overline{a}~r_3~\overline{a}~r_4~\overline{w}~\mathit{cxt}~\widehat{a}~\underline{b}~p_1~p_2~\widetilde{w}~p_3~p_4\]

\subsection{A  biological toy example}\label{subsec:toy}

\noindent
We consider a biological toy  example in the style of gene's alternative splicing \cite{WBB13}, 
where a gene $a$  codes for a protein $T$ when molecules 
$G$ is present and $C$ is absent, and in the opposite situation $a$ codes for protein $T'$.
This behavior is encoded in rules $1$ and $2$.
Then, rule 3 codes for the production of $C$ when proteins $T$ and $F$ are present, and $T'$ absent;
rule 4 codes for the production of $G$ when  proteins $T'$ is present and $F$ is absent.

>>>>

Here, for the entities $q$ and $w$ that are not provided by the context, we let:
\[
\begin{array}{lcl@{\hspace{1cm}}lcl}
P_q & \defeq & E(q,\widetilde{q}) & \overline{P}_q & \defeq & E(\overline{q},\widetilde{q}) \\
P_w& \defeq & E(w,\widetilde{w}) & \overline{P}_w & \defeq & E(\overline{w},\widetilde{w})
\end{array}
\]
where:
\[
\begin{array}{lcl}
E(q,\widetilde{q})&\defeq & \sum_{h=1}^3 (\startchain{q_i}\chainedlink{q_o}{\noact}\chainend{\noact})^h\ \link{\widetilde{q}_i}{\widetilde{q}_o}.P_q\\
&& +\\
&& \sum_{h=1}^3 (\startchain{q_i}\chainedlink{q_o}{\noact}\chainend{\noact})^h.\overline{P}_q\\
\end{array}
\]
In fact the presence/absence of $q$ and $w$ will be exploited by at least one rule and at most three rules.

@@

Now, we assume that we have an initial configuration containing the entities $q$ and $b$:
% (with $\widetilde{q},\widetilde{w},\widetilde{a},\widetilde{b}$ grouping all the possibile decorations for these names):
\[
\begin{array}{lcl}
\mathit{Sys}  \defeq \ %\restrict{r1,r2,r3,r4,\widetilde{q},\widetilde{w},\widetilde{a},\widetilde{b}}
 I\mid P_q \mid \overline{P}_w  \mid  \overline{P}_a  \mid P_b  \mid  P_1  \mid  P_2  \mid  P_3  \mid P_4  \mid \mathit{Cxt}.
\end{array}
\] 
Then, only the second reaction can be applied, and the transition carries the complete label $\upsilon$ below
{\tiny
\[
%\begin{array}{l}
%\upsilon = \\
%\restrict{r1,r2,r3,r4,\widetilde{q},\widetilde{w},\widetilde{a},\widetilde{b}}\\
 \startchain{\silent}
 \chainedlink{r_1}{r_1}
 \chainedlink{\overline{a}_i}{\bf  \overline{a}_i}\chainedlink{\bf \overline{a}_o}{\overline{a}_o}
 \chainedlink{r_2}{r_2}
 \chainedlink{q_i}{\bf q_i}\chainedlink{\bf q_o}{q_o}
 \chainedlink{b_i}{\bf b_i}\chainedlink{\bf b_o}{b_o}
 \chainedlink{\overline{w}_i}{\bf \overline{w}_i}\chainedlink{\bf \overline{w}_o}{\overline{w}_o}
 \chainedlink{\overline{a}_i}{\bf \overline{a}_i}\chainedlink{\bf \overline{a}_o}{\overline{a}_o}
 \chainedlink{r_3}{r_3}
 \chainedlink{\overline{a}_i}{\bf \overline{a}_i}\chainedlink{\bf \overline{a}_o}{\overline{a}_o}
 \chainedlink{r_4}{r_4}
 \chainedlink{\overline{w}_i}{\bf \overline{w}_i}\chainedlink{\bf \overline{w}_o}{\overline{w}_o}
 \chainedlink{\mathit{cxt}}{\mathit{cxt}}
 \chainedlink{\widehat{a}_i}{\bf \widehat{a}_i}\chainedlink{\bf \widehat{a}_o}{\widehat{a}_o}
 \chainedlink{\underline{b}_i}{\bf \underline{b}_i}\chainedlink{\bf \underline{b}_o}{\underline{b}_o}
 \chainedlink{p_1}{p_1}
 \chainedlink{p_2}{p_2}
 \chainedlink{\widetilde{w}_i}{\bf \widetilde{w}_i}\chainedlink{\bf \widetilde{w}_o}{\widetilde{w}_o}
 \chainedlink{p_3}{p_3}
 \chainedlink{p_4}{p_4}
 \chainend{\silent}
%\end{array}
\]}
The parts in bold are provided by the entity processes, the other parts are provided by the processes encoding the reactions and by the process encoding the context (starting at $\mathit{cxt}$ and ended at $p_1$.
In the label we can read that the rules $1$ and $4$ have been not executed because the entity $a$ is absent, 
the rule $3$ has been not applied because the entity $w$ is absent, then only rule $2$ has been applied, and it has produced the entity $w$. Also,  the context provides entity $a$, that will be available in the next state, and not the entity $b$.
Now, to let the label more readable, we show the result of the application of the function $flat(\cdot)$ to it:
\[
r_1~\overline{a}~r_2~q~b~\overline{w}~\overline{a}~r_3~\overline{a}~r_4~\overline{w}~\mathit{cxt}~\widehat{a}~\underline{b}~p_1~p_2~\widetilde{w}~p_3~p_4\]

\subsection{A  biological toy example of gene expression}\label{subsec:toy}

\noindent
We consider a biological toy  example in the style of gene's alternative splicing \cite{WBB13}.
Alternative splicing is a regulated process during gene expression that results in a single gene 
coding for multiple proteins. 
In practice, particular exons of a gene may be included within or excluded from the final, 
processed messenger RNA (mRNA) produced from that gene.

where a gene $a$  codes for a protein $T$ when molecules 
$G$ is present and $C$ is absent, and in the opposite situation $a$ codes for protein $T'$.
This behavior is encoded in rules $1$ and $2$.
Then, rule 3 codes for the production of $C$ when proteins $T$ and $F$ are present, and $T'$ absent;
rule 4 codes for the production of $G$ when  proteins $T'$ is present and $F$ is absent.

### 2020-04-05 06b-bisimulation.tex
modificati 7 files
modificati 7 files

The classical notion of bisimulation for process algebra equates two processes
when one process can simulate all the instructions executed by the other one and viceversa.
In its weak formulation, internal instructions, i.e. non visible by external observers,
are abstracted away. 
There are many variants of the bisimulation for process algebras,  for example the
barbed bisimulation~\cite{10.1007/3-540-55719-9_114} only consider the execution of invisible actions, and then equates two processes when the expose the same prefixes; for the mobile ambients~\cite{CardelliG00}, a process algebra equipped with a reduction semantics, a notion of
behavioural equivalence equates two processes when they expose the same ambients~\cite{GC03}.

>>>>

The classical notion of bisimulation for process algebras equates two processes
when one process can simulate all the instructions executed by the other one and viceversa.
In its weak formulation, internal instructions, i.e. non visible by external observers,
are abstracted away. 
There are many variants of the bisimulation for process algebras,  for example the
barbed bisimulation~\cite{10.1007/3-540-55719-9_114} only consider the execution of invisible actions, and then equates two processes when the expose the same prefixes; for the mobile ambients~\cite{CardelliG00}, a process algebra equipped with a reduction semantics, a notion of
behavioural equivalence equates two processes when they expose the same ambients~\cite{GC03}.
